---
title: "AN588-Week-5"
author: "Brooke Rothamer"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Module 09

Preliminary: install {curl}

## Introduction to Statistical Inference

In the last module, we learned standard errors are key to calculating confidence intervals and to basic inferential statistics.

In Module 7, we calculated confidence intervals for one of our estimates of a population parameter (the population mean, an estimand), based on a sample statistic (the sample mean, our estimator)

* The general way to define a confidence interval based on data from a sample is as the value of the statistic being considered (e.g., the mean) ± some critical value × the standard error of the statistic.

* The critical value is derived from the standardized version of a sampling distribution (e.g., the normal distribution) that corresponds the quantile limits we are interested in. For example, for the 95% confidence interval around the mean, the critical value corresponds the range of quantiles above and below which we expect to see only 5% of the distribution of statistic values. This is equivalent to the ± 1 - (α/2) quantiles, where α=0.05, i.e., the ± 0.975 quantile that we have used before for calculating 95% CIs.

* The standard error is the standard deviation of the sampling distribution, which, as noted above, is often estimated from the sample itself as σ/sqrt(n) but can also be calculated directly from the population standard deviation, if that is known.

* In Module 8, we created a vector, v, containing 1000 random numbers selected from a normal distribution with mean 3.5 and standard deviation 4. We then calculated the mean, standard deviation, and standard error of the mean (SEM) based on a sample of 30 observations drawn from that vector, and we used the normal distribution to characterize the quantiles associated with the central 95% of the distribution to define the upper and lower bounds of the 95% CI:
```{r mod 8 recap 1}
n <- 1000
mu <- 3.5
sigma <- 4
v <- rnorm(n, mu, sigma)
s <- sample(v, size = 30, replace = FALSE)
m <- mean(s)
m
```
``` {r mod 8 recap 2}
sd <- sd(s)
sd
```
``` {r mod 8 recap 3}
sem <- sd(s)/sqrt(length(s))
sem
```
``` {r mod 8 recap 4}
lower <- m - qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
upper <- m + qnorm(1 - 0.05/2) * sem  # (1-alpha)/2 each in the upper and lower tails of the distribution
ci <- c(lower, upper)
ci
```

## The Central Limit Theorem

The key importance of the **Central Limit Theorem** is that it states that the distribution of averages (or sums or other summary statistics…) of **iid** (independent and identically distributed) random variables becomes normal as the sample size increases. It is this fact that allows us to have a good sense of the mean and distribution of average events in a population even though we only observe one set of events and do not know what actual population distribution is. In fact, the CLT says nothing about the probability distribution for events in the original population, and that is exactly where its usefulness lies… that original probability distribution can be normal, skewed, all kinds of odd!

But we can nonetheless assume normality for the distribution of sample mean (or of the sum or mode, etc…) no matter what kind of probability distribution characterizes the initial population, *as long as our sample size is large enough and our samples are independent*. It is thus the CLT that allows us to make inferences about a population based on a sample.

To explore this a bit, let’s do some simulations. We are going to take lots of averages of samples from a particular non-normal distribution and then look at the distribution of those averages. Imagine we have some event that occurs in a population according to some probability mass function like the Poisson where we know λ=14. Recall, then, that the expectations of μ and σ2 for the Poisson distribution are both=λ.

Now let’s imagine taking a bunch of samples of size 10 from this population. We will take 1000 random samples of this size, calculate the average of each sample, and plot a histogram of those averages… it will be close to normal, and the standard deviation of the those averages - i.e., of the sampling distribution - should be roughly equal to the estimated standard error, the square root of (λ/n). [Recall that λ is the expected variance, so this is simply the square root of (expected variance / sample size)]

``` {r CLT 1}
#setting up the sample expectations
lambda <- 14
n <- 10
pop_se <- sqrt(lambda/n)  # the estimated SE
pop_se
```

``` {r CLT 2}
#Generating the samples and a histogram of the 1000 samples
x <- NULL
for (i in 1:1000) {
    x[i] <- mean(rpois(n = n, lambda = lambda))
}
hist(x, breaks = seq(from = lambda - 4 * sqrt(lambda)/sqrt(n), to = lambda +
    4 * sqrt(lambda)/sqrt(n), length.out = 20), probability = TRUE)
```

``` {r CLT 3}
#calculating the st dev of the samples
sd <- sd(x)
sd
#should be close to the pop_se
```

``` {r CLT 4}
qqnorm(x)
qqline(x)
```

Let’s do this for samples of size 100, too. We see that the mean stays the same, the distribution is still normal, but the standard deviation - the spread - of the sampling distribution is lower.

``` {r CLT 100 1}
#setting up the sample expectations
LAMBDA <- 14
N <- 100
POP_SE <- sqrt(LAMBDA/N)  # the estimated SE
POP_SE
```

``` {r CLT 100 2}
#Generating the samples and a histogram of the 1000 samples
X <- NULL
for (i in 1:1000) {
    X[i] <- mean(rpois(n = N, lambda = LAMBDA))
}
hist(X, breaks = seq(from = LAMBDA - 4 * sqrt(LAMBDA)/sqrt(N), to = LAMBDA +
    4 * sqrt(LAMBDA)/sqrt(N), length.out = 20), probability = TRUE)
```

``` {r CLT 100 3}
#calculating the st dev of the samples
SD <- sd(X)
SD
#should be close to the POP_SE
```

``` {r CLT 100 4}
qqnorm(X)
qqline(X)
```

We can convert these distributions to standard normals by subtracting off the expected population mean (λ) and dividing by the standard error of the mean (an estimate of the standard deviation of the sampling distribution) and then plotting a histogram of those values along with a normal curve.

``` {r CLT 100 5}
curve(dnorm(x, 0, 1), -4, 4, ylim = c(0, 0.8))
Z <- (X - LAMBDA)/POP_SE
hist(Z, breaks = seq(from = -4, to = 4, length.out = 20), probability = TRUE,
    add = TRUE)
```
Looks pretty normal.

Here’s an example of the CLT in action using sum() instead of mean()…

``` {r CLT 100 6}
n <- 100
x <- NULL
for (i in 1:1000) {
    x[i] <- sum(rpois(n = n, lambda = lambda))
}
hist(x, breaks = seq(min(x), max(x), length.out = 20), probability = TRUE)
```
Also pretty normal.

**Take Home Points:**

1. The CLT states that, regardless of the underlying distribution, the distribution of averages (or sums or standard deviations, etc…) based on a large number of independent, identically distributed variables:

  + will be approximately normal
  + will be centered at the population mean, and will have a standard deviation roughly equal to the standard error of the mean

* Additionally, variables that are expected to be the sum of multiple independent processes (e.g., measurement errors) will also have distributions that are nearly normal.

2. Taking the mean and adding and subtracting the relevant standard normal quantile × the standard error yields a confidence interval for the mean, which gets wider as the coverage increases and gets narrower with less variability or larger sample sizes.
 
3. As sample size increases, the standard error of the mean decreases and the distribution becomes more and more normal (i.e., has less skew and kurtosis, which are higher order moments of the distribution).

## Confidence Intervals for Sample Proportions

So far, we’ve talked about confidence intervals for sample means, but what about for other statistics, e.g., sample proportions for discrete binary variables. For example, if you have a sample of n trials where you record the success or failure of a binary event, you obtain an estimate of the proportion of successes, x/n. If you perform another n trials, the new estimate will vary in the same way that averages of a continuous random variable (e.g., zombie age) will vary from sample to sample. Taking a similar approach as above, we can generate confidence intervals for variability in the proportion of successes across trials.

Recall from our discussion of discrete random binary variables that the expectation for *proportion of successes*, which we will denote as **π** (where π, for “proportion”, is analogous to μ, for “mean”) is simply the *average number of successes across multiple trials*.

The expected sampling distribution for the proportion of successes is approximately normal centered at π and its standard deviation is estimated by sqrt(π(1−π)/n), which is, essentially, the standard error of the mean: it is the square root of (the expected variance / sample size). As above for μ, if we do not already have a population estimate for π, we can estimate this from a sample as p = x/n

Note: this expectation based on an approximation of the normal holds true only if both n×π and n×(1−π)
 are greater than roughly 5, so we should check this when working with proportion data.

### Challenge

Suppose a polling group in the Massachusetts is interested in the proportion of voting-age citizens in their state that already know they will vote for Elizabeth Warren in the upcoming November 5, 2024 general elections. The group obtains a yes or no answer from 1000 suitable randomly selected individuals. Of these individuals, 856 say they know they’ll vote for Senator Warren. 

How would we characterize the mean and variability associated with this proportion?
``` {r Mod 9 challenge 1}
n <- 1000 #sample size
x <- 856 #number of successes
phat <- x/n  # our estimate of pi
phat
```
Are n×π and n×(1−π) both > 5? Yes!
``` {r mod 9 challenge 2}
n * phat
```
``` {r mod 9 challenge 3}
n * (1 - phat)
```
``` {r mod 9 challenge 4}
pop_se <- sqrt((phat) * (1 - phat)/n)
pop_se
```
So, what is the 95% CI around our estimate of the proportion of people who already know how they will vote?
``` {r mod 9 challenge 5}
curve(dnorm(x, mean = phat, sd = pop_se), phat - 4 * pop_se, phat + 4 * pop_se)
upper <- phat + qnorm(0.975) * pop_se
lower <- phat - qnorm(0.975) * pop_se
ci <- c(lower, upper)
polygon(cbind(c(ci[1], seq(from = ci[1], to = ci[2], length.out = 1000), ci[2]),
    c(0, dnorm(seq(from = ci[1], to = ci[2], length.out = 1000), mean = phat,
        sd = pop_se), 0)), border = "black", col = "gray")
abline(v = ci)
abline(h = 0)
upper
lower
```

